Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-26 paper/paper.tex
section 2

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}

@@

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \label{fig:axioms}
  \caption{Kleene algebra axioms}
\end{figure}

The addition operation can be used to define the partial order
structure $a \leq b$ if $a + b = b$. In the theory of formal languages
this order structure can be used to model language containment. In this
section, we want to categorify the concept of Kleene algebra and
build on top of it in order to define an abstract theory of parsing.
We start by defining \emph{Kleene categories}.

@@

An unexpected example comes from the theory of substructural logics.

@@

Though Kleene categories are expressive enough to reason about concepts
that are outside of reach of Kleene algebras, their simply-typed nature
makes them not so expressive from a type theory point of view.

Since we are interested in design a type theory for reasoning about various
properties of parsers, we will assume that our Kleene category is defined
internal to a model of linear depedendently typed theory, which leads us
to the notion of models we are interested in:

\begin{definition}
  A model is a locally Cartesian category with \pedro{add the linear logic connectives here} equipped with an internal Kleene category.
\end{definition}

@@

\begin{figure}
  \label{fig:typewf}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %   
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %   
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
\end{figure}

@@

\begin{align*}
  c\tol c &\equiv I\\
  c\tol d &\equiv 0\\
  c\tol I &\equiv 0\\
  c\tol 0 &\equiv 0\\
  c\tol (A \otimes B) &\equiv (c\tol A) \otimes B + \nu(A) \otimes (c\tol B)\\
  c\tol A^* &\equiv (c\tol A) \otimes A^*\\
  c\tol (A + B) &\equiv (c\tol A) + (c\tol B)
\end{align*}

>>>>

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\nat}{\mathbb{N}}

@@

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \caption{Kleene algebra axioms}
  \label{fig:axioms}
\end{figure}

The addition operation can be used to define the partial order
structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
formal languages, this order structure can be used to model language
containment. In this section, we want to categorify the concept of
Kleene algebra and build on top of it in order to define an abstract
theory of parsing.  We start by defining \emph{Kleene categories}.

@@

This example illustrates how restricted Kleene algebras are in terms
of reasoning about languages. By only having at most one morphism
between objects, there is not a lot of information they can convey. In
this case, the only information you get is language containment. We
will see throughout this paper how the extra degrees of freedom
granted by having more morphisms give you more algebraic structure for
reasoning about languages.

For the next example, we see an unexpected connection with the theory
of substructural logics.

@@

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theory
point of view. This is limiting because type theories are tested
syntactic frameworks for manipulating complicated categories
structures while avoiding some issues common in category theory, such
as coherence issues.

With this in mind, we want to design a dependent type theory that
builds on top of Kleene categories with the goal of internalizing
various reasoning principles for languages and their parsers. This
leads us to the abstract notion of model we are interested in
capturing with our type theory.

\begin{definition}
  A model is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives, 
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

\begin{figure}
  \label{fig:typewf}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %   
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %   
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
\end{figure}

@@

\begin{align*}
  c\lto c &\equiv I\\
  c\lto d &\equiv 0\\
  c\lto I &\equiv 0\\
  c\lto 0 &\equiv 0\\
  c\lto (A \otimes B) &\equiv (c\lto A) \otimes B + \nu(A) \otimes (c\lto B)\\
  c\lto A^* &\equiv (c\lto A) \otimes A^*\\
  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
\end{align*}
